---
title: Comparing Automated Subcortical Volume Estimation Methods; Amygdala Volumes Estimated by FSL and FreeSurfer have Low Consistency.
author:
  - name: Patrick Sadil
    corresponding: true
    email: psadil1@jh.edu
    affiliations:
      - Johns Hopkins Bloomberg School of Public Health
  - name: [others TBD]
  - name: Martin Lindquist
    corresponding: false
    roles: []
    affiliations:
      - Johns Hopkins Bloomberg School of Public Health
keywords:
  - MRI
  - Amygdala
abstract: |
  Subcortical volumes are a promising source of biomarkers and features in biosignatures, and automated methods facilitate extracting them in large, phenotypically rich datasets. However, while extensive research has verified that the automated methods produce volumes that are similar to those generated by expert annotation, less research is available on the consistency of methods with each other. Using data from the UK Biobank, we compare the estimates of subcortical volumes produced by two popular sofware suites: FSL and FreeSurfer. Although most subcortical volumes exhibit good to excellent consistency across the methods, the tools produce diverging estimates of amygdalar volume. Through simulation, we show that this poor consistency can lead to conflicting results, where one but not the other tool suggests statistical significance, or where both tools suggest a significant relationship but in opposite directions. Considering these issues, we discuss several ways in which care should be taken when reporting on relationships involving amygdalar volume. 
plain-language-summary: |
  . ...
key-points:
  - Two popular methods of automated subcortical segmentation, FreeSurfer and FIRST, produce estimates of volume with variable consistency.
date: last-modified
bibliography: references.bib
number-sections: true
notebook-links: false
crossref:
  custom:
    - kind: float
      key: sfig
      latex-env: sfig
      reference-prefix: Figure A
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
      caption-prefix: Figure A
    - kind: float
      key: stbl
      latex-env: stbl
      reference-prefix: Table A
      space-before-numbering: false
      latex-list-of-description: Supplementary Table
      caption-prefix: Table A
---

```{r setup}
library(patchwork)
library(dplyr)
library(ggplot2)
library(agree)
library(tidyr)
library(purrr)

source(here::here("R/figures.R"))
targets::tar_load(ukb_vols_long)
targets::tar_load(ukb_vols) 
n_for_sim <- targets::tar_read(amyg_vols) |>
  dplyr::distinct(f.eid) |>
  dplyr::n_distinct()

```

# Introduction

Regional volumes of subcortex have been proposed as biomarkers for several psychopathologies. For example, the volume of the amygdala has been suggested as a biomarker for Alzheimer's, depression symptom severity in young adults, bipolar disorder in youth, migraine frequency, chronic pain, and others [@daftary_relationship_2019; @khatri_alzheimers_2022; @pfeifer_meta-analysis_2008; @liu_hippocampus_2017; @vachon-presseau_corticolimbic_2016; @rogers_smaller_2009; @ruocco_amygdala_2012; @szeszko_amygdala_2004]. As a biomarker, subcortical volumes are advantageous for being interpretable (given the rich literature linking these structures to many functions), explainable (hypotrophy and hypertrophy are both easily described to healthcare providers and patients), and readily available. The latter point comes from the fact that it is possible to estimate the regional volumes from any structural image with several automated algorithms.

For estimating regional subcortical volumes, two automated techniques are popular: FMRIB's Integrated Registration and Segmentation Tool (FIRST) from the FMRIB Software Library (FSL) and FreeSurfer's Automated Segmentation (ASEG) [@patenaude2007; @patenaude2011; @fischl2012]. Both techniques exhibit high consistency with the gold-standard of manual segmentation, in healthy adults and some clinical populations [@hsu2002; @tae2008; @morey2009; @pardoe2009; @dewey2010; @lehmann2010; @doring2011; @nugent_automated_2013; @wenger_comparing_2014], although there is variability across regions and between methods. For segmenting the hippocampus, FreeSurfer has been reported as having higher intraclass correlations than FSL [@doring2011], though neither method appears to have worse reliability (across repeated scans) than manual segmentation [@mulder_hippocampal_2014]. For the putamen, FSL has a higher Dice coefficient with manual segmentation, and the methods perform similarly on the caudate [@perlaki2017]. For the amygdala, which method performs better depends on the metric [@morey2009]. However, these comparisons may not generalize to other populations (e.g., pediatric, elderly), given that performance of the automated techniques is not consistently high across populations [@schoemaker2016; @kim2012; @sánchez-benavides2010; @zhou_charting_2021].

While comparisons to manual segmentation could identify which methods produce the best estimates of volume, it remains less clear how the two methods compare to each other. To our knowledge, the two methods have only been compared to each other by @perlaki2017. In their study, the methods appeared consistent with each other (exhibiting intraclass correlations that ranged from around 0.7 - 0.9), but the analyses included only the putamen and caudate, and the sample size was relatively small (30 participants). We extend the results of @perlaki2017 with the remaining structures and with a much larger population. Our investigation is in consideration of research that would use estimates of volume as a biomarker by, for example, correlating volume with a health-related outcome. Our primary concern is whether the results of such a study could be expected to depend on the method used for automated segmentation.

# Methods and Results

First, we looked at the consistency of subcortical volumes between FSL and FreeSurfer using data from the UK Biobank [@alfaro-almagro_image_2018]. The data were downloaded Jan 2024 and contained `r dplyr::n_distinct(ukb_vols$f.eid)` participants with usable anatomical data (Category 190: FreeSurfer ASEG; Category 1102: FSL FIRST). For details on intraclass correlation calculations, see [-@sec-icc-models]. Estimates and associated uncertainty are displayed using subscripts, as recommended by @louis_effective_2008. For example, an estimate of 0.22 with a 95% confidence interval spanning [0.21, 0.23] will be rendered as $_{0.21}0.22_{0.23}$.

:::{#fig-vol-comparison}

```{r fig-vol-comparison}
#| fig-height: 7

a <- ukb_vols |>
  plot_struct("Hippocampus", base_size = 8) |>
  GGally::ggmatrix_gtable() |>
  wrap_elements()
b <- ukb_vols |>
  plot_struct("Amygdala", base_size = 8) |>
  GGally::ggmatrix_gtable() |>
  wrap_elements()

a + b + 
  plot_layout(ncol = 1) + 
  plot_annotation(tag_levels = "a", tag_suffix = ")") & 
  theme(legend.position = "bottom")
```

Comparisons of Subcortical Volumes Estimated by FSL and FreeSurfer for two example regions a) Hippocampus and b) Amygdala. For remaining subcortical regions, see @stbl-iccs. In the lower triangular panels, the line of equivalence is marked with a solid line, and the dashed lines show the result of orthogonal regression. In the upper panel, the uncertainty estimates span 95% confidence intervals. The histograms along the diagonal display volumes in the full sample. 

:::

Across all structures, estimated agreement was lower than estimated consistency (@stbl-iccs), reflecting differences in the average volumes estimated by the two methods ([-@sec-avg-shift]). However, a constant shift across participants would not affect our primary focus, which is related to estimated correlations between regional volumes and some other measure. For that reason, we focus not on agreement but instead on consistency.

```{r contrasts}
#| eval: false
fit_amygdala <- lme4::lmer(
  Volume ~ Method*Hemisphere + (1 | f.eid), 
  filter(ukb_vols_long, Structure == "Amygdala") |> 
    mutate(f.eid = factor(f.eid)))

fit_amygdala_contrasts <- fit_amygdala |>
  emmeans::emmeans(pairwise ~ Hemisphere*Method) |>
  emmeans::contrast(interaction="pairwise")


```


Consistency varies by region (@fig-vol-comparison, @stbl-iccs). To interpret the intraclass correlations, consider the categories provided by @cicchetti1981: <0.4: poor, [0.4, 0.6): fair, [0.6, 0.75): good, [0.75, 1): excellent. Using those categories, the methods exhibit "good" to "excellent" consistency for most regions. However, the consistency of volumes for the amygdala is markedly worse than the others (@fig-vol-comparison$\~$b), being around only $_{0.24}0.24_{0.25}$ and $_{0.21}0.22_{0.23}$ for the left and right hemispheres (ranges span 95% confidence intervals). Compare the amygdala to the hippocampus (@fig-vol-comparison$\~$a), which has good consistency (left: $_{0.69}0.69_{0.7}$, right: $_{0.7}0.7_{0.71}$). For both structures (i.e., the amygdala and hippocampus), consistency across hemispheres as reported by FreeSurfer is the highest numerically (@fig-vol-comparison). For the amygdala, there is a two-way interaction between method and hemisphere (FSL - FreeSurfer: 0.0587, $p<0.001$), with FreeSurfer reporting that the right amygdala is larger than the left (left - right: 0.0351, $p<0.001$) and FSL reporting that the left is larger than the right (left - right: -0.1882, $p<0.001$).

The amygdala has been described as particularly challenging to segment, with one small study (N=23) reporting a consistency of 0.6 for volumes estimated by FSL across repeated scans of the same individual [@morey2010]. Moreover, automated segmentation algorithms can be affected by experimental factors like site, scanner, participant positioning, and software version [@hedges2022; @du2021; @mcguire2017; @yang2016; @liu2020; @mulder_hippocampal_2014; @morey2010; @perlaki2017], and differential sensitivity to such factors could impact an intraclass correlation. In the UKB, several potentially confounding factors were correlated with estimates of amygdalar volume (@sfig-partials-amyg). However, regressing these factors from the estimates of volume did not improve the consistency between the methods ([-@sec-confounds]).

With poor consistency between measurements of the amygdala, there is concern that reported relationships involving amygdala volumes may depend on which method is used for estimating the volume, a choice that may be considered arbitrary or lab-specific. At least two kinds of issues could arise. First, lower consistency could make it more likely that one but not both methods leads to significant correlations. Second, lower consistency could make it more likely that the two methods produce significant correlations that go in opposite directions. 

::: {#fig-sm-errors}

```{r fig-sm-errors}
#| fig-width: 8
#| fig-height: 6

targets::tar_load(c(icc_data2))
targets::tar_load(cog_dif_sig)
targets::tar_load(cog_dif_dir)

a <- icc_data2 |>
  filter(icc < 0.5) |>
  mutate(across(c(icc, rho), factor)) |>
  pivot_longer(
    c(starts_with("dd"), starts_with("ds")), 
    names_to = c("type", "level"), 
    names_sep = "_",
    values_to = "value") |>
  mutate(
    type = case_match(
      type, 
      "dd" ~ "Different\nDirection",
      "ds" ~ "Different\nSignificance"),
    Effect = factor(type, levels = c("Different\nSignificance", "Different\nDirection"))) |>
  pivot_wider(names_from = level) |>
  rename(ICC=icc) |>
  ggplot(aes(x=N, y=rho, fill=med)) +
  facet_grid(ICC~Effect, labeller = label_both) +
  geom_raster() +
  scale_fill_viridis_c(
    option="turbo",
    name="Estimated\nProportion",
    limits = c(0, 0.5)) +
  scale_x_continuous(
    "N Participants",
    breaks = c(0, 50, 100),
    labels = c(0, 50, 100)) +
  ggtitle("Artificial Data") +
  theme_gray(base_size = 9)


b <- left_join(
  cog_dif_sig |>
    select(predictor, N, ds_med, Hemisphere),
  cog_dif_dir |>
    mutate(estimate = abs(estimate)),
  by = join_by(predictor, N, Hemisphere)) |>
  filter(stringr::str_detect(Hemisphere, "Left")) |>
  na.omit() |>
  pivot_longer(
    c(starts_with("dd"), starts_with("ds")), 
    names_to = c("Effect", "level"), 
    names_sep = "_",
    values_to = "value") |>
  mutate(
    Effect = case_match(
      Effect, 
      "dd" ~ "Different\nDirection",
      "ds" ~ "Different\nSignificance"),
    Effect = factor(Effect, levels = c("Different\nSignificance", "Different\nDirection")),
    predictor = stringr::str_sub(predictor, 3),
    rho = glue::glue("({round(estimate, digits = 2)})"),
    Variable = interaction(predictor, rho, sep = " ", lex.order=FALSE)) |>
  ggplot(aes(x=N, y=Variable, fill=value)) +
  facet_wrap(~Effect, labeller = label_both, ncol=2) +
  geom_raster() +
  scale_fill_viridis_c(
    option="turbo",
    name = "Estimated\nProportion",
    limits = c(0,0.5)) +
  ylab("Cognitive Variable (rho)") +
  scale_x_continuous(
    "N Participants",
    breaks = c(0, 50, 100),
    labels = c(0, 50, 100)) +
  ggtitle("UKB")  +
  theme_gray(base_size = 9)

a + b + plot_layout(ncol=2, guides = "collect") + plot_annotation(tag_levels = "a", tag_suffix = ")") & theme(legend.position = "bottom")
```


Effects of Low Measurement Consistency. In the subfigures, the left column or panel shows the proportion of simulated experiments where one method produces a non-significant correlation and the other produces a correlation whose p-value is between 0.025 and 0.05. The right column or panel shows the proportion of simulated experiments where both measures are significantly correlated with an outcome but in opposite directions. Proportions are shown with color [@swihart_lasagna_2010]. a) Simulations with Artificial Data. Rows (rho) indicate pre-specified correlation. ICC: assumed intraclass correlation between measures of volume. b) Simulations with UKB. Each row corresponds to a non-image derived phenotype from the UKB. The value in parentheses is the absolute correlation of the variable with the average volumes of the left amygdala (average across methods). For a display without color that includes estimates of uncertainty, see @sfig-left-right.

:::

To investigate how often these two issues could occur, we first simulated experiments with artificial data. In each simulation, datasets with two noisy estimates of volume and a third, outcome, variable were generated such that the two volume estimates had a pre-specified intraclass correlation with each other and the true volume had a given product-moment correlation with the outcome variable. The estimated volumes were then tested for a product-moment correlation with the outcome, and the process was repeated for several intraclass correlations and sample sizes. In all simulations, the true product-moment correlation was set to a value that is small but non-zero (0.01), typical for neuroimaging research [0.1, @marek_reproducible_2022], or large (0.2). For additional details on the simulation methods, see [-@sec-sim-details].

With lower consistency, the estimated product-moment correlations were more often on opposite sides of a significance threshold (@fig-sm-errors$\~$a, left column). For example, with consistency near the value that was estimated for volumes of the amygdala in the UKB (0.2) and a typical effect size (0.1), significance differed in around $_{32.2}32_{32.4}$ percent of simulations (for simulations, ranges indicate 95% equal-tailed intervals). That percentage was only marginally improved by increasing the sample size to 100 (to $_{31.7}31.9\%_{32.2}$).

Lower consistency also coincided with a higher proportion of experiments in which the two methods correlate in opposite directions (@fig-sm-errors$\~$a, right column). Considering the previous example (an intraclass correlation of 0.2, an effect size of 0.1, and 50 participants per experiment), around $_{20.1}22\%_{23.3}$ of experiments in which both product-moment correlations were significant resulted in the correlations having opposite signs. As the intraclass correlation increased, the rates of this effect decreased rapidly, approaching 0% with an intraclass correlation of 0.4.

To assess how often these two issues could occur in practice, we repeated the above analyses but with data from the UKB. From the UKB, we extracted the "Cognitive" variables using the FMRIB UKBiobank Normalisation, Parsing And Cleaning Kit [@mccarthy_funpack_2023], restricting analyses to variables with instance 2 and that exhibited one of the largest (in absolute magnitude) 50 rank correlations (of the correlations between the cognitive variables and the average of the two estimates of the volume of the left amygdala), and further to only those participants that had values for all 50 of those variables (`r n_for_sim` participants). Each simulation resulted in a rank correlation between the two volume estimates for the left amygdala and each of the 50 variables (results were comparable with the right amygdala). For additional details, see [-@sec-sim-details].

Considering differing significance, the rates across experiments using the UKB matched the rates from experiments with artificial data. In simulated experiments with 50 UKB participants in which at least one correlation was significant, one correlation was not significant in between $_{25.5}25.6_{25.8}$ and $_{30.8}31.1_{31.3}$ percent of simulations (the range spans the 50 cognitive variables; @fig-sm-errors$\~$b, left panel). That percentage was generally lower for variables that were more strongly correlated with the measure of volume. For illustration, compare the variables that are higher versus lower in the panel.

Considering significant correlations with differing signs, the rates across experiments with the UKB were generally lower than the rates with artificial data--that is, there was more concordance between the methods. In experiments simulated with 50 UKB participants, the rates ranged from $_{1.3}1.4\%_{1.6}$ to $_{10.1}10.7\%_{11.4}$. For most variables, increasing the number of participants decreased the proportion of experiments in which the two correlations exhibited opposite signs, but for some the proportion increased. Across the 50 variables, 9 had a higher proportion at N=100 than N=50, 2 of which had non-overlapping 95% equal-tailed intervals (6348: duration to complete numeric path; 400: time to complete round of pairs matching game; see also @sfig-left-right).

# Discussion

We examined the consistency of subcortical volumes within the UK Biobank [@alfaro-almagro_image_2018], observing that two common methods of estimating the volume of the amygdala, one from FSL and one from FreeSurfer, have poor consistency with each other. Simulations revealed that consistency this low could often negatively impact the interpretability of results in two ways: the two methods may produce correlations that are on opposite sides of significance thresholds, and if they are both significant then they may be correlated in opposite directions. The prevalence of these occurrences was estimated with artificial data and data from the UKB.

## Recommendations

### When testing for new biomarkers, report relationships with multiple automated methods (e.g., both FSL and FreeSurfer).

The main concerns in this report are that two common methods for estimating amygdalar volume have poor consistency, and that this poor consistency can lead to conflicting results. Researchers may have idiosyncratic reasons for selecting a method, particularly when the choice is viewed as arbitrary. If the choice between methods is arbitrary, then reporting the outcome across selections clarifies the fragility or robustness of a result [for a general discussion, see @steegen_increasing_2016]. The choice may not always be arbitrary, as there are metrics and study populations for which one method may perform better [@zhou_charting_2021; @morey2009]. Note that one effect of low consistency could be a downward bias on the magnitude of estimated correlations (@sec-measurement-error), and so there may be advantages to not only reporting but also combining the estimates across methods when predicting health-related outcomes (e.g., by averaging, or including both as independent variables in predictive models). Reporting estimates from multiple reasonable tools will help move conclusions beyond "there exists a correlation with the volume of the amygdala as estimated by method M (version x)" to simply "there exists a correlation with the volume of the amygdala".


### When reviewing or conducting meta-analyses of relationships with amygdala volume, consider the method that was used to estimate volume.

As mentioned in the Introduction, the amygdala has received substantial attention for being predictive of an array of health-related outcomes. As presented in this report, the volume that is estimated by one automated method may only weakly correspond to the volume estimated by another, and so it may be misleading to conduct a meta-analysis without accounting for the algorithms used by the individual studies. In the UKB, there are differences in the strength of the correlations between the measures of volume and the cognitive variables, with the magnitude of the correlations involving FreeSurfer's method being numerically larger in almost all of the variables considered in this report (@sfig-pop-cor). Larger correlations do not imply higher veracity, but they further indicate that the methods track aspects of amygdalar anatomy differently.

### When replicating or extending research on a relationship that involves the volume of the amygdala, use the method reported in the original publications.

This recommendation follows standard practice for a replication study. We highlight it here in consideration of both extension studies that aim to apply a biomarker or biosignature that includes amygdala volume (such as when testing a putative relationship in a new population), and also in consideration of the ongoing evolution of methods for automatically estimating subcortical volumes. Although FSL and FreeSurfer are two of the most popular methods, others exist [e.g., @akhondi-asl2013], including newer techniques based on deep-learning approaches [e.g., @billot2023; for review, see @singh2021]. Newer methods may have better correspondence with manual segmentation, earning them consideration for replication or extension studies. But as this report shows, two methods can perform well while exhibiting poor consistency with each other. So when building on prior findings, it remains important to use the methods of those prior findings, even when they are superseded. 

# References {.unnumbered}

::: {#refs}
:::

\newpage
\appendix

# Supplementary Materials {#sec-supplements}

## Intraclass Correlation {#sec-icc-models}

The intraclass correlation was based on a two-way mixed linear model [@mcgraw1996]. In the model, the volume $x$ for the region of participant $i$ as measured by method $k$ was treated as equal to the sum of an intercept, $\nu$, the "true" volume, $\lambda$, a method bias, $c_k$, and an error term, $\epsilon$

$$
\begin{aligned}
x_{ik} &= \nu + \lambda_i + c_k + \epsilon_{ik} \\
\sum_k c_k &= 0 \\
\lambda_i &\sim N(0,\sigma_\lambda^2) \\
\epsilon_{ik} &\sim N(0, \sigma_\epsilon^2)
\end{aligned}
$$

Note that the $c_k$ terms are assumed fixed, with variance given by $\sigma_c^2=\sum_k c_k^2/(k-1)$.

An important assumption of this model is that the two methods are expected to have the same mean-squared error. The model does not include features that would allow for the errors in the two methods to be correlated (e.g., participants are not distinguished by characteristics that coincide with the methods performing better or worse).

The consistency version of the intraclass correlation, $ICC(C,1)$, and the absolute agreement, $ICC(A,1)$ were given as fractions of variance

$$
\begin{aligned}
ICC(C,1) & = \frac{\sigma_\lambda^2}{\sigma_\lambda^2 + \sigma_\epsilon^2} \\
ICC(A,1) & = \frac{\sigma_\lambda^2}{\sigma_\lambda^2 + \sigma_\epsilon^2 + \sigma_c^2}
\end{aligned}
$$

Models were estimated with `lme4` and intraclass correlations were calculated with `ReX` [@bates_fitting_2015; @xu_rex_2023].

## Differences in Average Volumes {#sec-avg-shift}

Although we primarily focused on the consistency of the methods, shifts in the average volumes reported by the two methods were observed (@sfig-biases-perc), patterns that have been previously observed [@gomez-ramirez2022; @perlaki2017; @dewey2010]. FSL tends to report volumes that are larger than those from FreeSurfer for the Accumbens, Amygdala, Hippocampus, and Pallidum, whereas for the Caudate, Putamen, and Thalamus FSL tends to report values that are smaller than those from FreeSurfer (all $p < 0.0001$ for two-sided t-test). Across structures, the variability in the difference appears to covary with the average of the two volume estimates, increasing as the average volume estimate decreases (@sfig-biases-perc$~$b). 

::: {#sfig-biases-perc}

```{r sfig-biases-perc}
#| fig-height: 7

a <- ukb_vols_long |>
  pivot_wider(names_from = Method, values_from = Volume) |>
  mutate(
    .average = (FIRST + FreeSurfer) / 2,
    .difference = (FIRST - FreeSurfer) 
  ) |>
  ggplot2::ggplot(ggplot2::aes(x = .average, y = .difference)) +
  facet_wrap(~Structure, scales = "free") +
  scattermore::geom_scattermore(alpha=0.1, pointsize = 3) +
  stat_density2d(
    aes(
      alpha = after_stat(ndensity), 
      fill = after_stat(ndensity)), 
    geom = "raster", 
    contour = FALSE) +
  geom_ba() +
  scale_fill_viridis_c(option = "inferno") +
  scale_alpha(range = c(0,1), guide = "none") +
  xlab("Average (FSL + FreeSurfer)/2") +
  ylab("Difference (FSL - FreeSurfer)") 

b <- ukb_vols_long |>
  pivot_wider(names_from = Method, values_from = Volume) |>
  mutate(
    .average = (FIRST + FreeSurfer) / 2,
    .difference = (FIRST - FreeSurfer) /.average * 100
  ) |>
  ggplot2::ggplot(ggplot2::aes(x = .average, y = .difference)) +
  scattermore::geom_scattermore(alpha=0.1, pointsize = 3) +
  stat_density2d(
    aes(
      alpha = after_stat(ndensity), 
      fill = after_stat(ndensity)), 
    geom = "raster", 
    contour = FALSE) +
  geom_ba() +
  scale_fill_viridis_c(option = "inferno") +
  scale_alpha(range = c(0,1), guide = "none") +
  xlab("Average (FSL + FreeSurfer)/2") +
  ylab("Difference (FSL - FreeSurfer)/Average (%)") 

a + b + 
  plot_layout(nrow = 2, guides = "collect") + 
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```

Bland-Altman Plots for Subcortical Volume. Horizontal lines show average difference and limits of agreement (1.96 standard deviations), with ribbons marking 95% confidence intervals. Panels correspond to subcortical structures. Volumes measured in cm$^3$. Left and right structures are plotted together. Shifts in the average estimate correspond to the central ribbon excluding zero. The color overlay indicates the degree of overplotting. a) Raw Differences. Note that axes are across panels independently. b) Differences by Percent Average.

:::


\newpage 
## Intraclass Correlations Across All Subcortical Regions {#sec-other-structs}

::: {#stbl-iccs}

```{r stbl-iccs}

ukb_vols_long |>
  filter(stringr::str_detect(Method, "FSSUBSEG", negate=TRUE)) |>
  pivot_wider(names_from = Method, values_from = Volume) |>
  group_nest(Hemisphere, Structure) |>
  mutate(
    icc = map(
      data,
      ~select(.x, FIRST, FreeSurfer) |>
        irr::icc(model="twoway")),
    lower = map_dbl(icc, pluck, "lbound"),
    upper = map_dbl(icc, pluck, "ubound"),
    icc = map_dbl(icc, pluck, "value"),
    across(c(icc, lower, upper), \(x) round(x, 2)),
    icc = glue::glue("@@{lower}@{icc}@@{upper}@"),
    iccA = map(
      data,
      ~select(.x, FIRST, FreeSurfer) |>
        irr::icc(model="twoway",type = "agreement")),
    lower = map_dbl(iccA, pluck, "lbound"),
    upper = map_dbl(iccA, pluck, "ubound"),
    iccA = map_dbl(iccA, pluck, "value"),
    across(c(iccA, lower, upper), \(x) round(x, 2)),
    iccA = glue::glue("@@{lower}@{iccA}@@{upper}@"),
  ) |>
  select(Structure, Hemisphere, `ICC(C,1)`=icc, `ICC(A,1)`=iccA) |>
  arrange(Structure) |>
  gt::gt() |> 
  gt::text_transform(
    fn = subscript_formatting, 
    locations = gt::cells_body(columns=c("ICC(C,1)", "ICC(A,1)"))
  ) |>
  gt::as_latex()
```

Intraclass Correlation for Subcortical Structures. Subscripts indicate 95% confidence intervals.

:::

## Intraclass Correlations After Residualizing on Potential Confounds {#sec-confounds}

One possible source of low intraclass correlations could be systematic inaccuracies with certain kinds of participants. For example, one of the two methods could tend to underestimate volumes when given brains that have experienced severe atrophy, which would decrease the agreement of the methods. To assess this, correlations between the amygdala volumes and the SIMPLE confounds within the UKB were calculated [@alfaro-almagro2021].

Several of the correlations appeared to be non-zero (@sfig-partials-amyg). However, regressing these confounds from the estimated amygdala volumes did not improve the intraclass correlations (@stbl-iccs-deconfounded).

::: {#sfig-partials-amyg}

```{r sfig-partials-amyg}
#| fig-height: 7

targets::tar_load(partials)

partials |>
  dplyr::select(
    Variable=Parameter1, 
    Summary=Parameter2, 
    lower=CI_low, 
    upper=CI_high, 
    Hemisphere) |>
  mutate(Variable = forcats::fct_relabel(Variable, stringr::str_replace_all, "_" , ":")) |>
  ggplot(aes(y=Variable, color=Summary)) + 
  geom_errorbarh(aes(xmin=lower, xmax=upper)) + 
  facet_wrap(~Hemisphere, labeller = label_both) +
  xlab("Product-Moment Correlation with Summarized Amygdala Volumes") +
  theme_gray(base_size = 9) +
  theme(legend.position = "bottom") 

```

Correlation Between Potential Confounders and Amygdala Volume Measurements. Summary describes how the the volumes were combined across methods. Summaries were either an average or a difference (FSL-FreeSurfer). Note that the variable "Head Size" corresponds to a scaling factor, and that larger values imply smaller brains.

:::

::: {#stbl-iccs-deconfounded}

```{r stbl-iccs-deconfounded}

ukb_vols |>
  mutate(acq_day = (acq_day - min(acq_day)) |> as.numeric()) |>
  na.omit() |>
  mutate(
    age2 = age^2,
    acq_day2 = acq_day^2,
    across(
      c(age, age2, t1_motion, head_size, acq_day, acq_day2), 
      \(x) (x - median(x)) / mad(x)),
    .by = site) |>
  pivot_longer(
    c(contains("Left"), contains("Right")),
    names_to = c("Structure", "Method"),
    names_pattern = "([[:print:]]+)_(FIRST|FreeSurfer)",
    values_to = "Volume") |>
  mutate(
    m = mean(Volume),
    Volume = resid(lm(Volume ~ site + (age*sex):site + age2:site + t1_motion:site + head_size:site + acq_day:site + acq_day2:site)),
    Volume = Volume + m,
    .by = c(Structure, Method)) |>
  select(-m) |>
  pivot_wider(names_from = Method, values_from = Volume) |>
  separate(Structure, c("Hemisphere", "Structure")) |>
  group_nest(Hemisphere, Structure) |>
  mutate(
    icc = map(
      data,
      ~select(.x, FIRST, FreeSurfer) |>
        irr::icc(model="twoway")),
    lower = map_dbl(icc, pluck, "lbound"),
    upper = map_dbl(icc, pluck, "ubound"),
    icc = map_dbl(icc, pluck, "value"),
    across(c(icc, lower, upper), \(x) round(x, 2)),
    icc = glue::glue("@@{lower}@{icc}@@{upper}@"),
    iccA = map(
      data,
      ~select(.x, FIRST, FreeSurfer) |>
        irr::icc(model="twoway",type = "agreement")),
    lower = map_dbl(iccA, pluck, "lbound"),
    upper = map_dbl(iccA, pluck, "ubound"),
    iccA = map_dbl(iccA, pluck, "value"),
    across(c(iccA, lower, upper), \(x) round(x, 2)),
    iccA = glue::glue("@@{lower}@{iccA}@@{upper}@"),
  ) |>
  select(Structure, Hemisphere, `ICC(C,1)`=icc, `ICC(A,1)`=iccA) |>
  arrange(Structure) |>
  gt::gt() |> 
  gt::text_transform(
    fn = subscript_formatting, 
    locations = gt::cells_body(columns=c("ICC(C,1)", "ICC(A,1)"))
  )|>
  gt::as_latex()

```

Intraclass Correlation After Deconfounding. Prior to calculating consistency, volumes were deconfounded by a version of the SIMPLE parameter set described by @alfaro-almagro2021. For the list of variables, see @sfig-partials-amyg. Subscripts indicate 95% confidence intervals.

:::

### Adjusting for ICV {#sec-adjust-icv}

When reporting differences in volume between groups, it is common to adjust for either head size or cerebral volume [@barnes_head_2010; @mathalon_correction_1993; @voevodskaya_effects_2014]. Adjusting by intracranial volume as estimated by FreeSurfer did not improve the intraclass correlations (@stbl-adjust).

::: {#stbl-adjust}


```{r stbl-adjust}
#| fig-pos: 't'

ukb_vols_long |>
  filter(stringr::str_detect(Method, "FSSUBSEG", negate=TRUE)) |>
  pivot_wider(names_from = Method, values_from = Volume) |>
  mutate(
    across(
      c(ends_with("FreeSurfer"), ends_with("FIRST")), 
      \(x) mean(x) + resid(lm(x ~ icv)))) |>
  group_nest(Hemisphere, Structure) |>
  mutate(
    icc = map(
      data,
      ~select(.x, FIRST, FreeSurfer) |>
        irr::icc(model="twoway")),
    lower = map_dbl(icc, pluck, "lbound"),
    upper = map_dbl(icc, pluck, "ubound"),
    icc = map_dbl(icc, pluck, "value"),
    across(c(icc, lower, upper), \(x) round(x, 2)),
    icc = glue::glue("@@{lower}@{icc}@@{upper}@"),
    iccA = map(
      data,
      ~select(.x, FIRST, FreeSurfer) |>
        irr::icc(model="twoway",type = "agreement")),
    lower = map_dbl(iccA, pluck, "lbound"),
    upper = map_dbl(iccA, pluck, "ubound"),
    iccA = map_dbl(iccA, pluck, "value"),
    across(c(iccA, lower, upper), \(x) round(x, 2)),
    iccA = glue::glue("@@{lower}@{iccA}@@{upper}@"),
  ) |>
  select(Structure, Hemisphere, `ICC(C,1)`=icc, `ICC(A,1)`=iccA) |>
  arrange(Structure) |>
  gt::gt() |> 
  gt::text_transform(
    fn = subscript_formatting, 
    locations = gt::cells_body(columns=c("ICC(C,1)", "ICC(A,1)"))
  ) |>
  gt::as_latex()

```

Consistency of Volumes when Adjusting by Intracranial Volume. Adjustments were done by residualizing with respect to ICV. Subscripts indicate 95% confidence intervals.

:::

## Simulated Experiments {#sec-sim-details}

To simulate hypothetical data, the models described in the previous sections were used. Sample sizes were set between 10 to 100 in steps of 10. At each combination of parameters, experiments were repeated 1,000,000 times. Simulated experiments with the UKB data were performed analogously to those with hypothetical data (UKB samples were taken with replacement).

Across repetitions, the rates of each effect were estimated by analytic Bayesian methods (binomial likelihood with Beta prior whose shape parameters were set to 1/2). The effect of "Different Significance" was calculated as the median of the posterior for which one correlation exhibited a $p$-value that was between 0.025 and 0.05 and the other was above 0.05, out of experiments in which at least one $p$-value was below 0.05. The effect of "Different Direction" was calculated as the median of the posterior, using experiments in which both correlations had opposite magnitudes given that both were significant. 

The 95% equal-tailed interval of the posteriors for the simulations are shown in @sfig-left-right.

::: {#sfig-left-right}

```{r sfig-left-right}
#| fig-height: 11
#| fig-width: 8

a <- make_spaghetti(cog_dif_sig, cog_dif_dir, "Left")
b <- make_spaghetti(cog_dif_sig, cog_dif_dir, "Right")

a + b + 
  plot_layout(guides = "collect", ncol = 1) + 
  plot_annotation(tag_levels = "a", tag_suffix = ")") &
  theme(legend.position = "bottom")

```

Effects of Low Measurement Consistency on the UKB in Left and Right Hemispheres. Ribbons span 95% equal-tailed interval estimated from simulated experiments. See also @fig-sm-errors, where the medians are represented with color.

:::


## Measurement Error and Reduced Correlation Magnitude {#sec-measurement-error}

Using the model from [-@sec-icc-models], we can build a measurement noise model [e.g., @frost_correcting_2000]. Call the final target for which we hope to find a relationship with the volume $y$. The relationship between that value and the volume was given by an ordinary linear regression with error $\delta$.

$$
\begin{aligned}
y_i & = \beta_0 + \beta_1\lambda_i + \delta_i \\
\delta_i & \sim N(0, \sigma_\delta^2) \\
\end{aligned}
$$

But since we do not know $\lambda$, the volumes estimated by the tools are used instead, changing the regression coefficient

$$
y_i = \beta_0 + \tilde{\beta_1}x_i + \delta_i \\
$$

Dilution occurs because the coefficient $\tilde{\beta_1}$ estimated in this model tends to be closer to zero, decreased by a factor related to the intraclass correlation [@frost_correcting_2000].

$$
\tilde{\beta_1} = \beta_1\frac{\sigma_\lambda^2}{\sigma_\lambda^2 + \sigma_\epsilon^2}
$$

Correspondingly, the desired correlation, $\rho=cor(y,\lambda)$, will also be biased.

$$
\begin{aligned}
\beta_1 & = \rho\frac{\sigma_\delta}{\sigma_\lambda} \\
\tilde{\beta_1} & = \tilde{\rho}\frac{\sigma_\delta}{sd(x)} \\ 
& =  \tilde{\rho} \frac{\sigma_\delta}{\sqrt{\sigma_\epsilon^2 + \sigma_\lambda^2}} \\
&  \implies \\
\rho \frac{\sigma_\delta}{\sigma_\lambda} &  = \tilde{\rho}\frac{\sigma_\delta}{\sqrt{\sigma_\epsilon^2+\sigma_\lambda^2}} \frac{\sigma_\lambda^2 + \sigma_\epsilon^2}{\sigma_\lambda^2} \\
&  \implies \\
\rho & = \tilde{\rho} \frac{\sqrt{\sigma_\epsilon^2+\sigma_\lambda^2}}{\sigma_\lambda }
\end{aligned}
$$

## Correlations Between Volumes of the Amygdala and Cognitive Variables

::: {#sfig-pop-cor}

```{r sfig-pop-cor}
#| fig-height: 7

targets::tar_load(amyg_vols_cors)

amyg_vols_cors |> 
  summarise(
    lower=quantile(estimate, probs = 0.025), 
    upper=quantile(estimate, probs = 0.975), 
    .by = c(Method, predictor, Hemisphere)) |> 
  left_join(
    distinct(cog_dif_dir, rho=estimate, predictor, Hemisphere)) |> 
  mutate(
    predictor = interaction(predictor, round(abs(rho), digits=2)),
    predictor = ordered(predictor), 
    predictor = forcats::fct_relabel(predictor, ~stringr::str_extract(.x, "(?<=f.)([[:print:]]+)(?=.0.[[:digit:]]+$)")),
    Method = stringr::str_replace(Method, "FIRST", "FSL")) |> 
  ggplot(aes(y=predictor, color=Method)) + 
  facet_wrap(~Hemisphere, scales = "free_y", labeller = label_both) +
  geom_errorbarh(aes(xmin=lower, xmax=upper)) +
  ylab("Cognitive Variable") +
  xlab("Rank Correlation with\nAmygdala Volume") +
  theme_gray(base_size = 9) +
  theme(
    legend.position = "bottom"
  )

```

Correlations Between Cognitive Factors and Estimated Volumes of the Left and Right Amygdala. Variables are ordered by decreasing rank correlation using average of left hemisphere volume estimates. Note that variables were selected based on the magnitude of their correlation with amygdalar volumes, which differed across hemispheres, and so the variables in left and right panels differ. Error bars span 95% confidence intervals (bootstrapped with 1000 samples).

:::
